Tidy data structure to support exploration and modeling of temporal-context data

### 2018-08-14 main.Rmd
added explanations on longitudinal data

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntactical and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a succinct and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The architecture of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

## Time series and longitudinal data

Time series and longitudinal data are the common terms when referring to temporal-context data. Despite of exactly the same data input, its representation diverges in these two areas due to different modelling focuses. Time series can be univariate or multivariate, but usually involves observations of relatively large in length possessed by each series. Time series researchers and analysts who are concerned with this large $T$ property, are mostly interested in modelling temporal dependency. Time series are represented as vectors or matrices in statistical software, for example MATLAB [@Matlab] and R [@R-base] (including `ts`, **zoo** [@zoo2005], and **xts** [@R-xts]). In the matrix case, each row is assumed to contain observations measured at a time point and each column to contain a single time series. This implies that data columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric cases. In addition to homogeneity, time indices must be stripped off the data and implicitly inferred as attributes or meta-information; the number of observations must be the same across all the series. Data wrangling from "raw" form to this specialist format can be frustrating and difficult, in order to perform a variety of downstream tasks such as analytics.

On the other hand, those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm]. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less restrictions to be made. [R]

>>>>

---
title: "Tidy data structure to support exploration and modeling of temporal-context data"
author:
  - familyname: Wang
    othernames: Earo
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : earo.wang@monash.edu
    correspondingauthor: true
  - familyname: Cook
    othernames: Dianne
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : dicook@monash.edu
  - familyname: Hyndman
    othernames: Rob J
    address: Department of Econometrics and Business Statistics, \newline Monash University, VIC 3800 \newline Australia.
    email : rob.hyndman@monash.edu
abstract: 'Temporal-context data is often rich with information and time formats, for example multiple observational units, different time lengths, heterogeneous data types, nested and crossed factors and etc. This work presents a cohesive and conceptual framework for organizing and manipulating temporal data, which in turn flows into visualization and forecasting routines. Tidy data principles are applied and extended to temporal data: (1) mapping the semantics of a dataset into its physical layout, (2) an explicitly declared index variable representing time, (3) a "key" comprised of single or multiple variables to uniquely identify units over time, using a syntax-based and user-oriented approach in which it imposes nested or crossed structures on the data. This tidy data representation most naturally supports thinking of operations on the data as building blocks, forming part of a "data pipeline" in time-based context. A sound pipeline practice facilitates a succinct and transparent workflow for analyzing temporal data. Applications are included to illustrate tidy temporal data structure, data pipeline ideas and usage. The infrastructure of tidy temporal data has been implemented in the R package **tsibble**.'
keywords: "temporal context, time series, data structure, R"
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: false
cover: true
toc: false
bibliography: [references.bib, rpkgs.bib]
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: no
    number_sections: yes
    citation_package: biblatex
---

@@

## Representing time series and longitudinal data

Time series and longitudinal data are the common terms when referring to temporal-context data. Despite of exactly the same data input, its representation diverges in these two areas due to different modelling focuses. Time series can be univariate or multivariate, but usually involves observations of relatively large in length (i.e. large $T$) possessed by each series. Time series researchers and analysts who are concerned with this large $T$ property, are mostly concerned with stochastic process, for the primary purpose of forecasting, temporal dynamics, and etc. Time series are represented as vectors or matrices in statistical software, for example MATLAB [@Matlab] and R [@R-base] (including `ts`, **zoo** [@zoo2005], and **xts** [@R-xts]). The structure the multivariate time series are in is called **wide format** where each row is assumed to hold observations at a time point and each column to contain a single time series. This implies that data columns of homogeneous types: numeric or non-numeric, but there are limited supporting methods for non-numeric cases. In addition to homogeneity, time indices must be stripped off the data and implicitly inferred as attributes or meta-information; the number of observations must be the same across all the series. Data wrangling from "raw" form to this specialist format can be frustrating and difficult, in order to perform a variety of downstream tasks such as analytics.

On the other hand, those who are engaged in longitudinal design, are primarily interested in explaining variations among individuals in change process and making inference about the underlying population. Longitudinal data or panel data typically refers to fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs to longitudinal design that some individuals are not observed at certain points in time, resulting in an unbalanced panel. This data issue leads representing longitudinal data much more commonly in **long format** where each row holds an observational unit per time and each column contains measurements across the individuals for all points in time. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less restrictions to be made. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm].
